---
title: "Rate of sexual debut, marriage, and marital dissolutions in Sub-Saharan Africa"
author:
  - name: Văn Kính Nguyễn
    id: vkn
    email: kinh.nguyen@uni-heidelberg.de
    affiliation:
      - ref: 1
  - name: Jeffrey Imai-Eaton
    affiliation:
      - ref: 2
affiliations: 
  - id: 1
    department: Heidelberg Institute for Global Health
    name: Heidelberg University Hospital
    city: Heidelberg
    state: Baden-Würtenburg
    country: Germanty
    address: Im Neuenheimer Feld 130/3
    postcode: 69120
    url: uni-heidelberg.de
  - id: 2
    department: Imperial College London
    name: Heidelberg University Hospital
    city: Heidelberg
    state: Baden-Würtenburg
    country: Germanty
    address: Im Neuenheimer Feld 130/3
    postcode: 69120
    url: uni-heidelberg.de

execute: 
  cache: true
  warning: false
  error: false
  include: false
  echo: false

editor:
  render-on-save: true

format:
  elsevier-pdf:
    journal:
      name: "medArxiv"
      formatting: preprint
      model: 3p
      layout: twocolumn
    number-sections: false
    mainfont: "Libertinus Serif"
    sansfont: "Libertinus Sans"
    monofont: "Libertinus Mono"
    include-before-body:
      text: |
        \linenumbers
    include-in-header:
      text: |
        \usepackage{amsmath}
        \setlength{\parindent}{0pt}
        %\usepackage[affil-it]{authblk}
        %\renewcommand{\Affilfont}{\footnotesize}
        \usepackage{hyperref}
        \hypersetup{colorlinks, linkcolor={blue}, citecolor={blue}, urlcolor={blue}}
        \usepackage[capitalise,noabbrev]{cleveref}
        \usepackage[title]{appendix}
        \usepackage[inline]{enumitem}
        %\usepackage{paralist}
        \usepackage{blkarray} % named matrixed
        \usepackage{lineno}
        \usepackage{etoolbox}
        \AtBeginEnvironment{abstract}{\setlength{\parindent}{0pt}}
        %\usepackage[hpos=0.15\paperwidth,vpos=0.1\paperheight,colormodel=RGB,color={[RGB]{253, 246, 227}}]{draftwatermark}
        %\SetWatermarkText{draft}
        %\SetWatermarkScale{1}
        %\usepackage{xcolor}
        \definecolor{pageDark}{rgb}{0.992,0.9647,0.8902}
        \definecolor{textDark}{rgb}{0.396,0.482,0.514}
        \definecolor{pageLight}{rgb}{0.0,0.1686,0.2117}
        \definecolor{textLight}{rgb}{0.5137,0.5804,0.5882}
        \pagecolor{pageDark}\color{textDark} 
        %\pagecolor{pageLight}\color{textLight} 
    # linestretch: 1
    pagestyle: headings
    fontsize: 11pt
    colorlinks: true
    indent: false
    keep-tex: true

bibliography: "/Users/knguyen/zotero.bib"
latex-clean: true
fig-dpi: 300
abstract: > 
  **Background** 
  This paper details a continuous-time multistate model designed to analyze
  transitions through various life course stages, including sexual debut,
  marriage, marital dissolution (separation, divorce, widowhood), and
  remarriage. The model incorporates age-dependent transition intensities and
  allows for the inclusion of other covariates. We discuss the theoretical
  underpinnings, model specification, calculation of transition matrices, and
  the likelihood framework, particularly focusing on an approach suitable for
  panel data where exact transition times for all states may not be observed.
---

# Background

Tireless efforts in HIV/AIDS research and treatment have led to better life for
people living with HIV (PLHIV). Treatment is yet still a lifelong process, and
every year, new infections, while have been subsided since 1996, still amount to
more than a million cases globally [@unaids2023]. The new infections are
disproportional distributed among women of young age [@unaids2023], the period
during which multiple sexual related life events occur including sexual debut,
marriage, and marriage dissolution.

In the marital status categories, the HIV prevalence has been shown to be
higher. Remarried individuals form a large portion of the population and have a
higher-than-average HIV prevalence [@de_walque_association_2012], whereas widows
and widowers had exceptionally high HIV prevalence [@mcgrath_older_2009]. The
risk of acquiring HIV increased in remarriage couples [@maphosa_unmasking_2014]
and was significantly associated with being married or in one of the marriage
dissolution states [@kimani_marital_2013]. 

Social and cultural reasons could lead to the increasing risk and prevalence in
those groups. HIV-positive surviving spouses might not want to disclose their
status and lessen the precautionary measures as children could be expected from
the new marriages. Widowhood could lead to high-risk partnerships to be formed
when widowed men and women reconnect to the sexual network
[@adetunji_hivaids_2001;@lopman_role_2009]. A higher rate of partner acquisition
could be expected due to polygamous marriage, frequency of remarriage and
marriage dissolution. Those with multiple marriages have a higher probability of
HIV infection among women with circumcision spouses [@poulin_inquiry_2011]. A
widespread cultural practice in sub-Saharan Africa that could also contribute to
risk of HIV transmission is widow inheritance [@agot_widow_2010].

The population of HIV-positive partners, surviving spouses, and divorcee could
face more difficulties to to access supporting services. In many African
countries, divorce is seen as a sign of failure and social ostracism often
follows which can lead to marginalization. Often lack of education and have
limited incomes, divorced and widowed women could be more likely to engage in
transactional sex or multiple concurrent partnerships and unable to negotiate
safe sex practices with partners. They might also not have access to prevention
services because they do not have the legal rights required for accessing these
services without male support. Unaddressed, these individuals would not only
face a higher risk of life-threatening conditions but also fuels the continuing
of the epidemic. 

Joint actions by governments and international organizations are needed to
better understand the gap and address the needs of different marital status
populations. In order to understand the extends of the resources needed as well
as potential impacts of programs targeting the population, estimating of the
life-course transitions between the marital statuses are needed. Improved
estimates of the transition rates will inform the estimate of the target
population size and dynamic models for epidemic modelling and assessments.

# Materials & Methods

We used a multistate survival framework to model the rates of transition between
marital statuses. Multistate models (MSMs) is a robust framework for analyzing
the movement through a series of discrete states over time by specifying the
instantaneous risk of moving from state $i$ to state $j$ at time $t$. To
characterize temporal trends and geographic variation in the rate, we allowed
the parameters to vary by country and survey years. Changes over time were
represented by a second-order auto-regressive (AR2). The model was fitted
separately for male and female.

## Household survey data

Data on age at first sex (AFS), age at marriage (AAM), and current marital
status were obtained from Demographic and Health Surveys (DHS) in 37 countries.
Only nationally representative surveys with individual records available for
analysis were used. For AFS, respondents stated their age at first sexual
intercourse; valid responses were 'never,' an integer age, or 'at first
marriage/union,' which was substituted with the reported age at first marriage.
Respondents also recalled their age at first marriage or union. Recorded current
marital statuses were: "Never married nor lived in a consensual union," "Legally
or formally married," "Not legally or formally married but living with a
man/woman in a consensual union," "Widowed from a marriage or consensual union
and not remarried or not in a consensual union," "Divorced from a legal or
formal marriage and not remarried or in a consensual union," "Separated from a
marriage or consensual union," and "Married or living in a union."

## States transition

We defined seven states of transition from virgin from birth ($V$) to current
marital status sexually debuted (and not married) ($X$), married or in a union
($M$), three states of marriage dissolution including separated ($S$), divorced
($D$), widowed ($W$), and remarried or in a union again ($R$). A remarried event
is defined if the respondent reported that there had been more than one union
and currently is married, in a union, or is in one of the three stages of
marriage dissolution. 

Transitions are unidirectional for initial life stages and allow for cycles in
later marital statuses: V $\to$ X, V $\to$ M (direct marriage implies debut), X
$\to$ M, M $\to$ S, M $\to$ D, M $\to$ W, S $\to$ D, S $\to$ R, D $\to$ R, W
$\to$ R, R $\to$ S, R $\to$ D, R $\to$ W. The states S, D, W, and R can form
cycles. Furthermore, when the AFS is reported equal or greater than AAM we set
the transition as the respondent moved from a virgin directly to married. There
can be as many number of unions as theoretically allowed to traverse the paths
in those records with more than once unions. However, while marital dissolution
states are not observed, we assumed if a marriage dissolute within a year after
married, there was no intermediate states in between, except the current status 
is remarried.

<!-- # Log-skew-logistic baseline hazard 

We previously reparameterized of a skew-logistic for the survival models of age
at sexual debut, which showed a better fit than other commonly used distributions
[@nguyen2022]. This distribution is described elsewhere as a Generalized
Logistic Distribution, Dagum distribution, or Burr Type III distribution. We
parameterized the distribution by three parameters, $\lambda, p, \gamma$, where
$\lambda > 0$ is the scale, $p > 0$ the shape, and $\gamma > 0$ the
skewness parameter. The distribution is implemented as R functions `dskewlogis`,
`pskewlogis` in [kinh-nguyen/ktools](https:://github.com/kinh-nguyen/ktools).
The baseline hazard function for the distribution is: 
$$
\lambda_{rs0}(t) = \frac{\gamma p u}{t (1+u) ((1+u)^\gamma - 1)}
$$
where $u = (\lambda t)^{-p}$. The hazard rate $\lambda_{rs}(t)$ is:
\begin{equation}
\label{eq:dagum_ph_hazard} \lambda_{rs}(t | \mathbf{X}) =
\lambda_{rs0}(t; \lambda, p, \gamma) \exp(\boldsymbol{\beta}^T \mathbf{X})
\end{equation}
where $\mathbf{X}$ is the vector of covariates and
$\boldsymbol{\beta}$ is the vector of their coefficients. 
The hazard represents the instantaneous transition intensity from state $i$ to
state $j$ at age $t$ and a log-linear model of covariates $\mathbf{X}$  -->

<!-- # Exponentially decaying Log-logistic hazard 

We combines a Log-logistic base hazard with a modifying factor $1/(1+\delta
e^{kt})$ that introduces an exponential-like decay Log-logistic (EDLL), with the
hazard defined as:
$$h(t; \lambda, \beta, \delta, k) = \frac{\beta t^{\beta-1}}{(\lambda + t^\beta)(1+\delta e^{kt})}$$
where $\lambda = \alpha^\beta > 0$, with $\beta > 0$ is the shape parameter,
$\delta = e^{kT} > 0$ (with $T$ being the reflection point), and $k > 0$ is
the steepness parameter of the decay.
The four-parameter hazard can model phenomena where a log-logistic hazard
pattern is progressively attenuated, potentially leading to a (large) non-zero
fraction of long-term survivors or insusceptibility. The hazard can capture
non-monotonic risk with large reduction which can be beneficial is analysis of
large waning risk, treatment efficacy fade-out, or effective long-term
interventions. Example of the hazard function is shown in @fig-exampleHazard.
For numerical computation, the log-transformed hazard function $\ln(h(t))$ is
calculated as follows:
\begin{align*}
\ln(h(t)) &= \ln(\beta) + (\beta-1)\ln(t) \\
          &\quad - \ln(\lambda + t^\beta) - \ln(1+\delta e^{kt})
\end{align*}
where the last two terms are computed using log-sum-exponential. -->

<!-- # Inverse Gamma hazard

Hazard of transition from state $i$ to state $j$ at age $t$ is modeled as an
inverse gamma's hazard where its probability density function is
$$f(x; \mu, \sigma) = \frac{\beta^{\alpha}}{\Gamma(\alpha)} x^{-(\alpha+1)} e^{-\beta/x}$$
in which $\mu>0$, $\sigma>0$, $\alpha = 1/\sigma^2$ and 
parameter $\beta = \mu (\alpha + 1)$ [@rigby2020]. The 
survival function is
$$S(x; \mu, \sigma) = \frac{\gamma(\alpha, \beta/x)}{\Gamma(\alpha)}$$
where $\gamma(s,z) = \int_0^z u^{s-1}e^{-u}du$ is the lower
incomplete gamma function. The log-hazard function is then 
\begin{align*}
& \log h(x; \mu, \sigma) = \log(f(x; \mu, \sigma)) - \log(S(x; \mu, \sigma)) \\
& = \alpha \log \beta - (\alpha+1) \log x - \frac{\beta}{x} 
 - \log \gamma\left(\alpha, \frac{\beta}{x}\right)
\end{align*} -->

# Log-sinh-arcsinh hazard

Since empirical data shows a variety of shapes in the transition probability
over age, we selected the sinh-arcsinh distribution [@jones2009] and reparameterized it
to have a support on $(0, \infty)$, namely a log-sinh-arcsinh distribution
(LSAS). This is done by substituting $y = \log x$ and performed a change of
variable, i.e.,
$$
f(x) = \phi(r) c x^{-1} \sigma^{-1} (1 + z^2)^{-0.5} 
$$
where $z = (\log x - \mu)\sigma^{-1}$, 
$e_1 = \exp[\tau\cdot \text{asinh}(z)]$,
$e_2 = \exp[-\nu \cdot \text{asinh}(z)]$,
$r = 0.5 \left[ e_1 - e_2 \right]$,
$c = 0.5 \left[ \tau \cdot e_1 + \nu \cdot e_2 \right]$,
and $\phi(r)$ is the standard normal PDF evaluated at $r$.  Only $\mu$ is on
the real line, all $\sigma$, $\tau$, $\nu$ are positive parameters. The
cummulative distribution function is given by $F(x) = \Phi(r)$ and the hazard
function is $h(x) = f(x)/(1 - F(x))$. Implementation of the hazard function is
done in TMB code available on [kinh-nguyen/msm](https://github.com/kinh-nguyen/msm/).

## Transition intensities and probabilities matrix

The transition rates between the states is described with the intensity matrix
$Q(t, \mathbf{X})$, 
\begin{equation}\label{mat:Q}
\scriptsize
Q = 
\begin{blockarray}{cccccccc}
V & X & M & S & D & W & R \\
\begin{block}{(ccccccc)c}
  q_{11} & \lambda_{12} & \lambda_{13} & 0 & 0 & 0 & 0 & V \\
  0 & q_{22} & \lambda_{23} & 0 & 0 & 0 & 0 & X \\
  0 & 0 & q_{33} & \lambda_{34} & \lambda_{35} & \lambda_{36} & 0 & M \\
  0 & 0 & 0 & q_{44} & 0 & 0 & \lambda_{47} & S \\
  0 & 0 & 0 & 0 & q_{55} & 0 & \lambda_{57} & D \\
  0 & 0 & 0 & 0 & 0 & q_{66} & \lambda_{67} & W \\
  0 & 0 & 0 & \lambda_{74} & \lambda_{75} & \lambda_{76} & q_{77} & R \\
\end{block}
\end{blockarray}
\end{equation}
The off-diagonal elements $q_{ij}(t, \mathbf{X}) = \lambda_{ij}(t, \mathbf{X})$
for $i \neq j$, representing the intensity of transition from state $i$ to state
$j$. The diagonal elements $q_{ii}(t, \mathbf{X}) = - \sum_{j \neq i}
\lambda_{ij}(t, \mathbf{X})$, representing the negative of the total rate of
leaving state $i$, e.g., 
$q_{11}(t) = -(\lambda_{V,SD}(t) + \lambda_{V,M1}(t))$.

The remarried state $R$ is separated from $M$ to model individuals who reported
having more than once union. While the rate of marital dissolution
could be different from those who had married only once ($M$) and those had
remarried ($R$), all the marriage dissolution events were not observed in the
data. Thus, we set the marital dissolution rates to be the same for both
marriage and remarried states $M$ and $R$, i.e., $q_{33} = q_{77},\ \lambda_{34}
= \lambda_{74},\ \lambda_{35} = \lambda_{75},\ \lambda_{36} = \lambda_{76}$. In
addition, the three remarried rates from either of the three marriage
dissolution states ($S, D, W$) are only informed by the number of individuals
who reported having more than once union and currently married --- it is unknown
from which state the individual had remarried. Thus, we constrained the three
rates to be equal of the parameters, i.e., $\lambda_{47} = \lambda_{57} =
\lambda_{67}$.

## Model likelihood

The likelihood is based on individuals' observed histories which includes
segments of history where the state occupancy is continuously known and segments
where only the start and end states of an interval are known. 
Let $t_0$ be the initial age of observation, $t_c$ be the current age at
observation, $t_d$ be the age at sexual debut, and $t_{m}$ be the age at
first marriage. 
The survival probability in state $i$ from $t_a$ to $t_b$ is $\exp(-H_i(t_a,
t_b, \mathbf{X}))$ [@van2016multi], where 
\begin{flalign*}
H_i(t_a, t_b, \mathbf{X}) &= \int_{t_a}^{t_b} q_{ii}(u, \mathbf{X}) du
%& = \int_{t_a}^{t_b} -\sum_{j \neq i} \lambda_{ij}(u, \mathbf{X}) du
%&= -\left[ A_i(\mathbf{X})(t_b-t_a) + B_i \frac{t_b^2-t_a^2}{2} \right] \\
\end{flalign*}
For currently virgin observations, the individual has remained in state V from
$t_0$ to $t_c$  with
$$L_V = \exp(-H_V(t_0, t_c, \mathbf{X})).$$
Those made sexual debut at $t_d$, the path is then V($t_0$) $\to$ V($t_d^-$)
$\to$ X($t_d$) $\to$ X($t_c$) with the likelihood contribution as
\begin{align*}
L_{X} = & \exp(-H_V(t_0, t_d, \mathbf{X})) \times \lambda_{V,X}(t_d, \mathbf{X}) \\
        \times & \exp(-H_{X}(t_d, t_c, \mathbf{X})).
\end{align*}
Those are currently in their first marriage at $t_c$, there are two
sub-paths: who made sexual debut before married with transition V $\to$ X $\to$
M ($t_d < t_{m}$) has 
\begin{align*}
L_{M}^{(1)} & = \exp(-H_V(t_0, t_d, \mathbf{X})) \times \lambda_{V,X}(t_d,
\mathbf{X}) \\
& \times \exp(-H_{X}(t_d, t_{m}, \mathbf{X})) \times \lambda_{X,M}(t_{m}, \mathbf{X}) \\
& \times \exp(-H_{M}(t_{m}, t_c, \mathbf{X})),
\end{align*}
whereas those have $t_d = t_{m}$ or unknown $t_d$ are assumed transits V $\to$
M has
\begin{align*}
L_{M}^{(2)} & = \exp(-H_V(t_0, t_{m}, \mathbf{X})) \times \lambda_{V,M}(t_{M}, \mathbf{X}) \\
  & \times \exp(-H_{M}(t_{m}, t_c, \mathbf{X})).
\end{align*}
For those who are currently separated (S), divorced (D), widowed (W), or
remarried (R) at $t_c$, there are two phases: likelihood of reaching M at
$t_{m}$ either $L_{M1}^{(1)}$ or $L_{M1}^{(2)}$ 
\begin{align*}
L_{M}^{(1)} = & \exp(-H_V(t_0, t_d, \mathbf{X})) \lambda_{V,X}(t_d, \mathbf{X})  \\
& \exp(-H_{X}(t_d, t_{m}, \mathbf{X})) \lambda_{X,M}(t_{m}, \mathbf{X}).\\
L_{M}^{(2)} = & \exp(-H_V(t_0, t_m, \mathbf{X})) \lambda_{V,M}(t_m, \mathbf{X}).
\end{align*}
and likelihood of being in state $C \in \{S, D, W, R\}$ at $t_c$, given in M
at $t_{m}$. 
$$L_C = L_{M}^{(*)} \times (P(t_{m}, t_c, \mathbf{X}))_{M, C}$$
where $P(\cdot)_{i,j}$ an element from the probability transition matrix P-matrix.
The term $P$ accounts for all possible sequences of transitions that could occur
between $t_m$ and $t_{c}$ starting in state M and ending in state C
[@van2016multi;@jackson2022].

The P-matrix contains elements $p_{ij}(t_1, t_2, \mathbf{X})$ representing the
probability that an individual in state $i$ at time $t_1$ will be in state $j$
at time $t_2$. The matrix is computed by solving the differential equations:
$$
\frac{d}{dt} P(t_1, t, \mathbf{X}) = P(t_1, t, \mathbf{X}) Q(t, \mathbf{X})
$$
with the initial condition $P(t_1, t_1, \mathbf{X}) = \mathbf{I}$ 
to obtain $P(t_1, t_2, \mathbf{X})$.
When $Q(t, \mathbf{X})$ is
constant over the interval $[t_1, t_2]$, then $P(t_1, t_2, \mathbf{X}) =
\text{expm}(Q \cdot (t_2 - t_1))$, where
`expm` is the matrix exponential [@van2016multi].
Here, since $Q(t, \mathbf{X})$ is age-dependent, $P(t_1, t_2, \mathbf{X})$ is
computed by dividing the interval $[t_1, t_2]$ into small subintervals $\Delta
t_k$ within which $Q(t_k, \mathbf{X})$ is assumed constant. In particular,
$$ P(t_1, t_2, \mathbf{X}) \approx \prod_{k} \text{expm}(Q(t_k, \mathbf{X})
\cdot \Delta t_k)$$
Here, we use $\Delta t_k = 1$ year, and $Q(t_k, \mathbf{X})$ is
evaluated at the start of the year $t_k$.

## Computational

Calculating the matrix exponential is an expensive operation
[@molerNineteenDubiousWays2003], thus, for each parameter set, we pre-computed
the values for all the unique combinations of individual profile and only
queried the values during the likelihood calculation. Since there were over a
million individual records, to reduce the size of the data for computational
purposes, the standard time-split data format (each row an episode) was not used
but indicators for the likelihood's terms in \cref{eq_likelihood} were
calculated for each individual (each row an individual). Furthermore,
individuals with similar life-course were further aggregated to group in which
the group counts was used to weight the likelihood in \cref{eq_likelihood}, thus
reducing the number of unique calculation of the likelihood.

The model and likelihood calculation were implemented with Template Model
Builder which can be accessed on Github's repository
[kinh-nguyen/multistage]{https://github.com/kinh-nguyen/multistage}. Parameter
estimate was done with empirical Bayes. The priors were chosen to shrink the
regression coefficients towards zero when they were only weakly informed by the
data. In particular, $\beta_{0,rs} \sim N(0, 0.5)$, $\beta_{1, rs}\sim N(0,
0.1)$ and $c_(\text{country})\sim N(0, 0.1)$.  We computed posterior medians and
95\% credible intervals (95\% CI) of the outcomes of interests, including the
rate, transition probability, and hazard by ages, using posterior samples of
3000 parameters.

# Results

```{r}
library(ktools)
# install.packages("expm")
# install.packages("TMB")
# install.packages("elementalist")
# install.packages("ggh4x")
# library("expm")
# remotes::install_github('kaskr/adcomp/TMB')
library("TMB")
library(ggh4x)
# remotes::install_github("teunbrand/elementalist")
library(elementalist)
k.legend <- function(x = .8, y = 1, dir = "horizontal", size = 0.25) {
    theme(
        legend.position = c(x, y),
        legend.direction = dir,
        legend.key.size = unit(size, "cm"),
        legend.title = element_text(size = 9),
        legend.box.background = element_rect_round(fill = "white", color = "grey90", radius = unit(0.2, "snpc"))
    )
}
```

```{r, loadata}
vroom::vroom(here("data/svw/AO2015DHS.csv"), col_select = -1) %>%
  mutate(
    A = case_when(A == "V" ~ 0, A == "X" ~ 1, A == "M" ~ 2, A == "S" ~ 3, A == "D" ~ 4, A == "W" ~ 5, A == "R" ~ 6),
    Z = case_when(Z == "V" ~ 0, Z == "X" ~ 1, Z == "M" ~ 2, Z == "S" ~ 3, Z == "D" ~ 4, Z == "W" ~ 5, Z == "R" ~ 6),
    # across(where(is.numeric), as.integer)
  ) %>% 
  mutate(
    fit = case_when(
      A != Z & (end - start == 1) & Z != 6 ~ 0,
      A == Z & (end - start == 1) ~ 1,
      otherwise ~ 2
    )
  ) %>%
  filter(cc == "AO", sex == "female", sv == 2015) %>%
  select(id, A, Z, start, end, tx, tm, n, fit) %>%
  mutate(
    dstart = as.double(start),
    istart = start,
  ) %>%
  filter(dstart > 0) -> tdt
```

```{r}
# predicting data by age for average duration
tdt %>%
  filter(A == 1 | A == 2) %>%
  group_by(start) %>%
  summarise(
    tx = mean(tx),
    tm = mean(tm),
    .groups = "drop"
  ) %>%
  bind_rows(
    anti_join(tibble(
      start = 1:(max(tdt$end) + 1),
      tx = 0,
      tm = 0
    ), ., by = "start")
  ) %>%
  arrange(start) %>%
  mutate(gap = lead(start) - start) %>%
  allot(sim_data)
```

```{r, prepmodelfit}
N_PAR = 7
data <- list()
data$n_age <- max(tdt$end) + 1 # include zero
data$prior_mu <- c(log(16), 2.5) # mid age in data
# https://github.com/kinh-nguyen/naomi/blob/60ad718ad0de201d8449c166d45138e0be2e6a5d/src/tmb.cpp#L254
data$prior_sigma <- c(0, 2.5)
data$prior_nu <- c(0, 2.5)
data$prior_tau <- c(0, 2.5)
# https://github.com/kinh-nguyen/naomi/blob/60ad718ad0de201d8449c166d45138e0be2e6a5d/src/tmb.cpp#L212
data$prior_coef <- c(0, 2.5)
data$sim_data = as.matrix(sim_data)
data <- modifyList(data, as.list(tdt))
n_iid <- tdt$id %>% unique %>% length # better pass as factor to DATA_FACTOR

data$id <- data$id - 1
init <- list(
  mu = rep(log(20), N_PAR),
  log_sigma = rep(log(.5), N_PAR),
  nu = rep(-.1, N_PAR),
  log_tau = rep(log(0.1), N_PAR),
  itc = rep(0, N_PAR),
  b_tx = 0,
  b_tm = 0,
  iid = rnorm(n_iid, 0, 1),
  log_iid = 0
)
```

```{r}
#| eval: false
library(TMB)
ktools::tmb_unload("msm")
TMB::compile("code/msm.cpp", flags = "-Wno-ignored-attributes")
base::dyn.load(TMB::dynlib("code/msm"))
invisible(TMB::config(tape.parallel = 0, DLL = "msm"))
TMB::openmp(20)
obj <- TMB::MakeADFun(
  data, init,
  map = list(
    mu = c(1, 2, 3, 4, 4, 4, 5) %>% factor(),
    log_sigma = c(1, 2, 3, 4, 4, 4, 5) %>% factor(),
    nu = c(1, 2, 3, 4, 4, 4, 5) %>% factor(),
    log_tau = c(1, 2, 3, 4, 4, 4, 5) %>% factor(),
    itc = c(1, 2, 3, 4, 4, 4, 5) %>% factor()
    # itc = c(1, 2, 3, 4, NA, NA, NA) %>% factor()
    # itc = rep(NA, N_PAR) %>% factor()
  ),
  random = "iid",
  DLL = "msm", silent = FALSE, LaplaceNonZeroGradient = 0
)

fit <- nlminb(
  obj$env$last.par.best, obj$fn, obj$gr,
  control = list(trace = 0, eval.max = 1000, abs.tol = 1e-12)
)
fit
fit$par %>% name2list()
rp <- obj$simulate(obj$env$last.par.best)

tibble(
  mu = rp$exp_mu,
  sigma = rp$sigma,
  nu = rp$nu,
  tau = rp$tau
)
plott(rp$PP[1,2,2:51])
plott(rp$PP[1,3,2:51])
plott(rp$PP[2,3,2:51])
dir.create(here('fit'), showWarnings = FALSE)
saveRDS(rp, here('fit/testMW.rds'))
```

We computed the 1-year transition probabilities by age for the different marital
statuses, $P(S_{t+1} = s | S_{t} = r)$, starting with virgin at birth, i.e., the
probabilities of transitioning from virgin to sexual debut or to married, and
from sexual debut to married. As these events are observed (i.e., respondents
reported an age when the event occurred), the fitted probabilities are presented
together with the empirical probabilities. The fitted probabilities of remaining
in the same state, the complimentary probability, are also shown.
@fig-fittedExample shows that the fitted probabilities are close to the
empirical estimates. 

```{r}
#| label: fig-fittedExample
#| fig-cap: Estimated and empirical conditional 1-year transition probabilities from currently virgin or currently sexually debuted by age. Only these transitions can be defined from survey data and have empirical estimates, which are showed as points with size relative to the population-at-risk in the previous age. Lines are the fitted probabilities from our model.
#| include: true
#| fig-env: "figure*"
#| fig-pos: "!ht"
#| fig-width: 7
#| fig-height: 4
#| out-width: "100%"
rp <- readRDS(here('fit/testMW.rds'))
str(rp)
P_yearly = rp$PP
P_cum <- array(NA, dim = dim(P_yearly))
P_cum[,,1] <- diag(7)
for (k in 2:dim(P_yearly)[3]) 
  P_cum[,,k] <- P_cum[,,k-1] %*% P_yearly[,,k]

pe = tibble(
  start = 0:51,
  VV = P_yearly[1,1,],
  VX = P_yearly[1,2,],
  VM = P_yearly[1,3,],
  XX = P_yearly[2,2,],
  XM = P_yearly[2,3,],
  MM = P_yearly[3,3,],
  MS = P_yearly[3,4,],
  MD = P_yearly[3,5,],
  MW = P_yearly[3,6,],
  MR = P_yearly[4,7,],
) %>% 
pivot_longer(-start)

tdt %>%
  filter(A < 2, Z <= 2) %>%
  group_by(A, start, tx) %>%
  summarise(n_risk = sum(n), .groups = "drop") %>%
  allot(at_risk)

tdt %>%
  filter(A < 2, Z <= 2) %>%
  group_by(A, Z, start, tx) %>%
  summarise(n_move = sum(n), .groups = "drop") %>%
  left_join(at_risk, by = c("A", "start", "tx")) %>%
  mutate(p_empirical = n_move / n_risk) %>% 
  mutate(
    A = case_when(A == 0 ~ 'V', A == 1 ~ 'X', A == 2 ~ "M", A == 3 ~ "S", A == 4 ~ "D", A == 5 ~ "W", A == 6 ~ 'R', otherwise ~ NA_character_),
    Z = case_when(Z == 0 ~ 'V', Z == 1 ~ 'X', Z == 2 ~ "M", Z == 3 ~ "S", Z == 4 ~ "D", Z == 5 ~ "W", Z == 6 ~ 'R', otherwise ~ NA_character_),
    name = paste0(A,Z)) %>% 
  full_join(pe, by = c("start", "name")) %>% 
allot(pd)

pd %>%
  mutate(group = case_when(
    A == "V" ~ 1,
    A == "X" ~ 2,
    A == "M" ~ 3,
  ), 
    p_empirical = ifelse(A == "M", NA_real_, p_empirical),
  ) %>%
  filter(A != "M") %>%
  filter(start > 0) %>% 
  ggplot(aes(start, p_empirical)) +
  geom_point(aes(size = n_risk, color = factor(name)), shape = 21) +
  geom_line(aes(y = value, color = factor(name))) +
  ggh4x::facet_grid2(
    cols = vars(group),
    scales = "free_y",
    labeller = labeller(group = function(x) {
      recode(x, `1` = "Currently virgin", `2` = "Current sexual debuted", `3` = "Currently married")
    })
  ) +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, .1)) +
  guides(size = FALSE) +
  labs(x = 'Age', y = 'Transition probability', color = 'Transition') +
  k.theme + k.legend(.925, .5, dir = "vertical")
```

::: {#ref}
:::

# Appendix

```{r}
#| label: fig-exampleHazard
#| fig-cap: Illustration of the Log-logistic base hazard with exponential decay (EDLL) hazard function for $\lambda = \exp(8)$, $\beta = 2.5$, $\delta = 0.01$, and $k = 0.1875$. The hazard is defined as $h(t; \lambda, \beta, \delta, k) = \frac{\beta t^{\beta-1}}{(\lambda + t^\beta)(1+\delta e^{kt})}$. 
#| include: true
#| fig-env: "figure*"
#| fig-pos: "!ht"
#| fig-width: 7
#| fig-height: 5
#| out-width: "80%"

lambda <- exp(8)
beta   <- 2.5
delta  <- 0.01
k_val  <- 0.1875 

# Log-hazard
# h(t) = (beta * t^(beta-1)) / ((lambda + t^beta) * (1 + delta * exp(k*t)))
log_h_t_calculator <- function(t, lambda, beta, delta, k) {
  logspace_add <- function(logx, logy) 
    pmax(logx, logy) + log1p(exp(-abs(logx - logy)))
  log_denom1 <- logspace_add(log(lambda), beta * log(t))
  log_denom2 <- logspace_add(0, log(delta) + k * t)
  log(beta) + (beta - 1) * log(t) - log_denom1 - log_denom2
}

log_h_t_calculator(1:50, lambda, beta, delta, k_val) %>%
  exp() %>%
  plot(1:50, .,
    type = "l",
    xlab = "Time (t)", ylab = "Hazard h(t)",
    main = "Simulated Hazard Function"
  )
```
