---
title: "Rate of sexual debut, marriage, and marital dissolutions in Sub-Saharan Africa"
author:
  - name: Văn Kính Nguyễn
    id: vkn
    email: kinh.nguyen@uni-heidelberg.de
    affiliation:
      - ref: 1
  - name: Jeffrey Imai-Eaton
    affiliation:
      - ref: 2
  - name: Till Baernighausen
    affiliation:
      - ref: 1
affiliations: 
  - id: 1
    department: Heidelberg Institute for Global Health
    name: Heidelberg University Hospital
    city: Heidelberg
    state: Baden-Würtenburg
    country: Germany
    address: Im Neuenheimer Feld 130/3
    postcode: 69120
    url: https://www.uni-heidelberg.de
  - id: 2
    department: Epidemiology
    name: Harvard T.H. Chan School of Public Health
    city: Boston
    state: MA
    country: USA
    address: 677 Huntington Ave
    postcode: 02115
    url: https://www.harvard.edu/

execute: 
  cache: true
  warning: false
  error: false
  include: false
  echo: false

editor:
  render-on-save: true

format:
  elsevier-pdf:
    journal:
      name: "medArxiv"
      formatting: preprint
      model: 3p
      layout: twocolumn
    number-sections: false
    mainfont: "Libertinus Serif"
    sansfont: "Libertinus Sans"
    monofont: "Libertinus Mono"
    include-before-body:
      text: |
        \linenumbers
    include-in-header:
      text: |
        \usepackage{amsmath}
        \setlength{\parindent}{0pt}
        %\usepackage[affil-it]{authblk}
        %\renewcommand{\Affilfont}{\footnotesize}
        \usepackage{hyperref}
        \hypersetup{colorlinks, linkcolor={blue}, citecolor={blue}, urlcolor={blue}}
        \usepackage[capitalise,noabbrev]{cleveref}
        \usepackage[title]{appendix}
        \usepackage[inline]{enumitem}
        %\usepackage{paralist}
        \usepackage{blkarray} % named matrixed
        \usepackage{lineno}
        \usepackage{etoolbox}
        \AtBeginEnvironment{abstract}{\setlength{\parindent}{0pt}}
        %\usepackage[hpos=0.15\paperwidth,vpos=0.1\paperheight,colormodel=RGB,color={[RGB]{253, 246, 227}}]{draftwatermark}
        %\SetWatermarkText{draft}
        %\SetWatermarkScale{1}
        %\usepackage{xcolor}
        \definecolor{pageDark}{rgb}{0.992,0.9647,0.8902}
        \definecolor{textDark}{rgb}{0.396,0.482,0.514}
        \definecolor{pageLight}{rgb}{0.0,0.1686,0.2117}
        \definecolor{textLight}{rgb}{0.5137,0.5804,0.5882}
        \pagecolor{pageDark}\color{textDark} 
        %\pagecolor{pageLight}\color{textLight} 
    # linestretch: 1
    pagestyle: headings
    fontsize: 11pt
    colorlinks: true
    indent: false
    keep-tex: true

bibliography: "/Users/knguyen/zotero.bib"
latex-clean: true
fig-dpi: 300
abstract: > 
  **Background** 
  This paper details a continuous-time multistate model designed to analyze
  transitions through various life course stages, including sexual debut,
  marriage, marital dissolution (separation, divorce, widowhood), and
  remarriage. The model incorporates age-dependent transition intensities and
  allows for the inclusion of other covariates. We discuss the theoretical
  underpinnings, model specification, calculation of transition matrices, and
  the likelihood framework, particularly focusing on an approach suitable for
  panel data where exact transition times for all states may not be observed.
---

# Background

Tireless efforts in HIV/AIDS research and treatment have led to better life for
people living with HIV (PLHIV). Treatment is yet still a lifelong process, and
every year, new infections, while have been subsided since 1996, still amount to
more than a million cases globally [@unaids2023]. The new infections are
disproportional distributed among women of young age [@unaids2023], the period
during which multiple sexual related life events occur including sexual debut,
marriage, and marriage dissolution.

In the marital status categories, the HIV prevalence has been shown to be
higher. Remarried individuals form a large portion of the population and have a
higher-than-average HIV prevalence [@de_walque_association_2012], whereas widows
and widowers had exceptionally high HIV prevalence [@mcgrath_older_2009]. The
risk of acquiring HIV increased in remarriage couples [@maphosa_unmasking_2014]
and was significantly associated with being married or in one of the marriage
dissolution states [@kimani_marital_2013]. 

Social and cultural reasons could lead to the increasing risk and prevalence in
those groups. HIV-positive surviving spouses might not want to disclose their
status and lessen the precautionary measures as children could be expected from
the new marriages. Widowhood could lead to high-risk partnerships to be formed
when widowed men and women reconnect to the sexual network
[@adetunji_hivaids_2001;@lopman_role_2009]. A higher rate of partner acquisition
could be expected due to polygamous marriage, frequency of remarriage and
marriage dissolution. Those with multiple marriages have a higher probability of
HIV infection among women with circumcision spouses [@poulin_inquiry_2011]. A
widespread cultural practice in sub-Saharan Africa that could also contribute to
risk of HIV transmission is widow inheritance [@agot_widow_2010].

The population of HIV-positive partners, surviving spouses, and divorcee could
face more difficulties to to access supporting services. In many African
countries, divorce is seen as a sign of failure and social ostracism often
follows which can lead to marginalization. Often lack of education and have
limited incomes, divorced and widowed women could be more likely to engage in
transactional sex or multiple concurrent partnerships and unable to negotiate
safe sex practices with partners. They might also not have access to prevention
services because they do not have the legal rights required for accessing these
services without male support. Unaddressed, these individuals would not only
face a higher risk of life-threatening conditions but also fuels the continuing
of the epidemic. 

Joint actions by governments and international organizations are needed to
better understand the gap and address the needs of different marital status
populations. In order to understand the extends of the resources needed as well
as potential impacts of programs targeting the population, estimating of the
life-course transitions between the marital statuses are needed. Improved
estimates of the transition rates will inform the estimate of the target
population size and dynamic models for epidemic modelling and assessments.

# Materials & Methods

\noindent We developed a parametric semi-Markov multistate survival model (MSM)
to estimate the transition rates between marital statuses since birth till the
age of interview. A reparameterized four parameters log-Sinh-arcsinh (lnSHASH)
baseline hazard was implemented to permit variations in the tails across
transitions. We estimated the model's parameters varied by country and sex. A
clock-forward approach is adopted to describe transitions that depend on time
since entering a state, modelling as a proportional hazard as the duration
increase. Respondents' age at interview were used to scale the hazard
distribution's variance. Details are as follows.

## Household survey data

\noindent Data on respondents's country, age, sex, age at first sex (AFS), age
at first marriage (AFM), and current marital status were obtained from
Demographic and Health Surveys (DHS) in 37 countries. Only nationally
representative surveys with individual records available for analysis were used.
For AFS, respondents stated their age at first sexual intercourse; valid
responses were 'never,' an integer age, or 'at first marriage/union,' which was
substituted with the reported age at first marriage. Respondents also recalled
their age at first marriage or union. Recorded current marital statuses were:
"Never married nor lived in a consensual union," "Legally or formally married,"
"Not legally or formally married but living with a man/woman in a consensual
union," "Widowed from a marriage or consensual union and not remarried or not in
a consensual union," "Divorced from a legal or formal marriage and not remarried
or in a consensual union," "Separated from a marriage or consensual union," and
"Married or living in a union." DHS's survey weights were firstly scaled by
$10^6$ and Kish's formular [@kish1995]. We splitted the individual history since
birth to 1-year interval, resulting $>5$ million rows data.

## States transition

\noindent We defined seven states of transition from virgin from birth ($V$) to currently sexually debuted and not married ($X$), married or in a union ($M$), three states of marriage dissolution including separated ($S$), divorced ($D$), widowed ($W$), and remarried or in a union again ($R$). A remarried event is defined if the respondent reported that there had been more than one union and currently in states M, D, S, W. Transitions are unidirectional for initial life stages and allow for cycles in later marital statuses: V $\to$ {X, M}, X $\to$ M, M $\to$ {S, D, W} $\to$ R, R $\to$ {S, D, W}. From the AFM, the events D, S, W, R are interval censored, and there can be numerous traverse the paths in those with more than once unions. We assumed if a marriage dissolute within a year after married, there was no intermediate states in between, except the current status is remarried.

## Transition hazard model

\noindent The time-varying transition rates between the states is described with
the time-varying intensity matrix $Q(t, \mathbf{X})$, 
\begin{equation}\label{mat:Q}
\scriptsize
Q = 
\begin{blockarray}{cccccccc}
V & X & M & S & D & W & R \\
\begin{block}{(ccccccc)c}
  q_{11} & \lambda_{12} & \lambda_{13} & 0 & 0 & 0 & 0 & V \\
  0 & q_{22} & \lambda_{23} & 0 & 0 & 0 & 0 & X \\
  0 & 0 & q_{33} & \lambda_{34} & \lambda_{35} & \lambda_{36} & 0 & M \\
  0 & 0 & 0 & q_{44} & 0 & 0 & \lambda_{47} & S \\
  0 & 0 & 0 & 0 & q_{55} & 0 & \lambda_{57} & D \\
  0 & 0 & 0 & 0 & 0 & q_{66} & \lambda_{67} & W \\
  0 & 0 & 0 & \lambda_{74} & \lambda_{75} & \lambda_{76} & q_{77} & R \\
\end{block}
\end{blockarray}
\end{equation} 
The off-diagonal elements $q_{ij}(t, \mathbf{X}) = \lambda_{ij}(t, \mathbf{X})$, 
$i \neq j$, representing the intensity of transition from state $i$ to state
$j$. The diagonal elements $q_{ii}(t, \mathbf{X}) = - \sum_{j \neq i}
\lambda_{ij}(t, \mathbf{X})$, representing the negative of the total rate of
leaving.
The remarried state $R$ describes individuals who reported to having more than
once marriages/unions. Because the marriage dissolution events were not observed
in data, we constrained the rates of dissolution to be the same for ($M$) and
($R$), i.e., $q_{33} = q_{77},\ \lambda_{34} = \lambda_{74},\ \lambda_{35} =
\lambda_{75},\ \lambda_{36} = \lambda_{76}$. Also, the remarried rates from the
marriage dissolution states ($S, D, W$) are only informed by the number of
individuals with more than once union and currently married without knowing from
which state the individual had remarried. Thus, we constrained $\lambda_{47} =
\lambda_{57} = \lambda_{67}$. Among those who had only once or none unions, we
evaluated only the submatrix block $Q_{6\times 6}$ excluding state R.

The general transition intensities between state $i$ and $j$ over ages $t =
1,2,...,$ is
\begin{flalign*}
\lambda_{ij}(t) &= h_{ij}(t | \Theta) \exp(\alpha_{ij}) \\
\lambda_{23}(t) &= h_{23}(t | \Theta) \exp(\alpha_{23} + \beta_{23} t_x)  \\
\lambda_{3j}(t) &= h_{3j}(t | \Theta) \exp(\alpha_{3j} + \beta_{3j} t_m)
\end{flalign*}
where $h(t)$ is a parametric hazard function, $t_x, t_m$ are time since sexual
debut and time since first marriage, respectively. In other words, this is a
semi-Markov model where a transition depends not only on current state, but also 
proportionally to the time spent since entering a state.

Exploring the empirical risk estimate, we found different shapes in the
transition probability by age with a variety of dynamics after the peak, we opted for the flexible sinh-arcsinh distribution [@jones2009] and reparameterized to have a positive support. The density function reads as
$$ f(x|\mu, \sigma, \nu, \tau) =
\phi(r) c x^{-1} \sigma^{-1} (1 + z^2)^{-0.5}
$$ 
where 
$z = (\log x - \mu)\sigma^{-1}\tau^{-1}$, 
$\gamma =  \tau \log(z + (1 + z^2)^{-.5}) - \nu$, $r = \sinh(\gamma)$, 
$c = \cosh(\gamma)$, 
and $\phi(r)$ is the standard normal probability distribution function. The
parameter $\mu$ is the mean, with variance $\sigma > 0$, kurtosis $\tau > 0$,
skewness $\nu > 0$, i.e, model righ-skewed empirical data. The hazard function
is given by 
$h(x) = f(x)/(1 - \Phi(r))$. 
For each sex ($s$), and across countries ($k$), the four parameters are assumed
to vary around a global mean, e.g., $\mu_{ij,s,k} = \mu_{ij,s} + c_{ij,s,k}$,
$c_{ij, s,k}\sim N(0, \kappa_{ij, s})$. In addition, as older paricipants have
naturally a wider permited age range to report (i.e., the right tail of the
distribution), we modelled the kurtosis as a function of age at interview, i.e.,
$$\tau_{ij,sk} = \exp (\zeta_{ij,s,k} + \zeta_{ij,s,k} \cdot \text{aai})$$
in which *aai* is the standardized age at interview. 

## Model likelihood

\noindent The likelihood is calculated based on individuals' observed histories
which includes segments where the state occupancy is continuously known and
segments where only the start and end states of an interval are known. 
Assuming the hazard is constant in a 1-year episode, we can approximate the
transition probability that an individual in state $i$ at time $t$ will be in
state $j$ at time $t + 1$ by
$$\dfrac{d}{dt} P(t, t + 1, \mathbf{X}, \Theta) \approx \text{expm}(Q(t, X, \Theta)).$$
where `expm` is the matrix exponential function [@vandenhout2020].
For interval censored observations, those that transited away from state $M$ at
time $t$ to state $j$ at time $t + k$, the likelihood is a total probability of all 
possible paths traverse through every single 1-year steps, i.e.,
$$P(t, t + k) \approx \prod_{i = 0}^k \text{expm}(Q(t + i))$$
with $Q(t_k)$ is evaluated at the start of the interval. The likelihood is then the 
corresponding matrix's element obsevations.

The model's parameters was estimated using empirical Bayes approach. The
standard normal prior were chosen for the unrestricted parameters, and the
half-normal prior was used for positive parameters sampling on logarith scale.
We computed posterior medians and 95% uncertainty intervals (95% UI) of the
outcomes of interests, including the rate, transition probability, and hazard by
ages, using posterior samples of 3000 parameters. The model and likelihood
calculation were implemented with Template Model Builder which can be accessed
on Github's repository [kinh-nguyen/msm](https://github.com/kinh-nguyen/msm). 

# Results

```{r}
library(ktools)
# install.packages("expm")
# install.packages("TMB")
# install.packages("elementalist")
# install.packages("ggh4x")
# library("expm")
# remotes::install_github('kaskr/adcomp/TMB')
library("TMB")
library(ggh4x)
# remotes::install_github("teunbrand/elementalist")
library(elementalist)
k.legend <- function(x = .8, y = 1, dir = "horizontal", size = 0.25) {
    theme(
        legend.position = c(x, y),
        legend.direction = dir,
        legend.key.size = unit(size, "cm"),
        legend.title = element_text(size = 9),
        legend.box.background = element_rect_round(fill = "white", color = "grey90", radius = unit(0.2, "snpc"))
    )
}
```

```{r, loadata}
vroom::vroom(here("data/new_dw.csv.bz2")) %>%
  mutate(
    A = case_when(A == "V" ~ 0, A == "X" ~ 1, A == "M" ~ 2, A == "S" ~ 3, A == "D" ~ 4, A == "W" ~ 5, A == "R" ~ 6),
    Z = case_when(Z == "V" ~ 0, Z == "X" ~ 1, Z == "M" ~ 2, Z == "S" ~ 3, Z == "D" ~ 4, Z == "W" ~ 5, Z == "R" ~ 6),
    # across(where(is.numeric), as.integer)
  ) %>%
  mutate(
    fit = case_when(
      A != Z & (end - start == 1) & Z != 6 ~ 0,
      A == Z & (end - start == 1) ~ 1,
      otherwise ~ 2
    )
  ) %>%
  allot(full_d)

full_d %>%
  filter(cc == "CG") %>%
  select(A, Z, aai, start, end, tx, tm, n, fit) %>%
  mutate(
    dstart = as.double(start),
    istart = start,
    aai = scale(aai)
  ) %>%
  filter(dstart > 0) %>%
  filter(Z < 3) %>%
  allot(tdt)
```

```{r}
# predicting data by age for average duration
tdt %>%
  filter(A == 1 | A == 2) %>%
  group_by(start) %>%
  summarise(
    tx = mean(tx),
    .groups = "drop"
  ) %>%
  bind_rows(
    anti_join(tibble(
      start = 1:(max(tdt$end) + 1),
      tx = 0
    ), ., by = "start")
  ) %>%
  arrange(start) %>%
  mutate(gap = lead(start) - start) %>%
  allot(sim_data)
```

```{r, prepmodelfit}
N_PAR = 3
data <- list()
data$n_age <- max(tdt$end) + 1 # include zero
data$sim_data = as.matrix(sim_data)

data <- modifyList(data, as.list(tdt))

init <- list(
  mu = c(2.6, 2.8, 2.9),
  log_sigma = c(-2, -2, .1),
  log_nu = c(.5, -2, -1),
  log_tau = c(-2, -.5, .8),
  tau_aai = c(-.04, -.05, -.2),
  b_tx = c(-0.03),
  itc = c(3, -.5, 2)
)
str(init)
```

```{r}
#| eval: false
library(TMB)
ktools::tmb_unload("msm")
TMB::compile("code/msm.cpp", flags = "-Wno-ignored-attributes")
base::dyn.load(TMB::dynlib("code/msm"))
invisible(TMB::config(tape.parallel = 1, DLL = "msm"))
TMB::openmp(20)

obj <- TMB::MakeADFun(
  data, init,
  DLL = "msm", silent = FALSE, LaplaceNonZeroGradient = 0
)

fit <- nlminb(
  obj$env$last.par.best, obj$fn, obj$gr,
  control = list(trace = 0, eval.max = 1000, abs.tol = 1e-12, iter.max = 1000)
)
fit
fit$par %>% name2list()
rp <- obj$simulate(obj$env$last.par.best)
plott(rp$PP[1,2,2:51],width = 70)
plott(rp$PP[1,3,2:51],width = 70)
plott(rp$PP[2,3,2:51],width = 70)
dir.create(here('fit'), showWarnings = FALSE)
saveRDS(rp, here('fit/testMW.rds'))
```

We computed the 1-year transition probabilities by age for the different marital
statuses, $P(S_{t+1} = s | S_{t} = r)$, starting with virgin at birth, i.e., the
probabilities of transitioning from virgin to sexual debut or to married, and
from sexual debut to married. As these events are observed (i.e., respondents
reported an age when the event occurred), the fitted probabilities are presented
together with the empirical probabilities. The fitted probabilities of remaining
in the same state, the complimentary probability, are also shown.
@fig-fittedExample shows that the fitted probabilities are close to the
empirical estimates. 

```{r}
#| label: fig-fittedExample
#| fig-cap: Estimated and empirical conditional 1-year transition probabilities from currently virgin or currently sexually debuted by age. Only these transitions can be defined from survey data and have empirical estimates, which are showed as points with size relative to the population-at-risk in the previous age. Lines are the fitted probabilities from our model.
#| include: true
#| fig-env: "figure*"
#| fig-pos: "!ht"
#| fig-width: 7
#| fig-height: 4
#| out-width: "100%"
rp <- readRDS(here('fit/testMW.rds'))
str(rp)
P_yearly = rp$PP

pe = tibble(
  start = 1:dim(P_yearly)[3],
  VV = P_yearly[1,1,],
  VX = P_yearly[1,2,],
  VM = P_yearly[1,3,],
  XX = P_yearly[2,2,],
  XM = P_yearly[2,3,],
) %>% 
pivot_longer(-start)

tdt %>%
  filter(A < 2, Z <= 2) %>%
  group_by(A, start, tx) %>%
  summarise(n_risk = sum(n), .groups = "drop") %>%
  allot(at_risk)

tdt %>%
  filter(A < 2, Z <= 2) %>%
  group_by(A, Z, start, tx) %>%
  summarise(n_move = sum(n), .groups = "drop") %>%
  left_join(at_risk, by = c("A", "start", "tx")) %>%
  mutate(p_empirical = n_move / n_risk) %>% 
  mutate(
    A = case_when(A == 0 ~ 'V', A == 1 ~ 'X', A == 2 ~ "M", A == 3 ~ "S", A == 4 ~ "D", A == 5 ~ "W", A == 6 ~ 'R', otherwise ~ NA_character_),
    Z = case_when(Z == 0 ~ 'V', Z == 1 ~ 'X', Z == 2 ~ "M", Z == 3 ~ "S", Z == 4 ~ "D", Z == 5 ~ "W", Z == 6 ~ 'R', otherwise ~ NA_character_),
    name = paste0(A,Z)) %>% 
  full_join(pe, by = c("start", "name")) %>% 
allot(pd)

pd %>%
  mutate(group = case_when(
    A == "V" ~ 1,
    A == "X" ~ 2,
    A == "M" ~ 3,
  ), 
    p_empirical = ifelse(A == "M", NA_real_, p_empirical),
  ) %>%
  filter(A != "M") %>%
  filter(start > 0) %>% 
  ggplot(aes(start, p_empirical)) +
  geom_point(aes(size = n_risk, color = factor(name)), shape = 21) +
  geom_line(aes(y = value, color = factor(name))) +
  ggh4x::facet_grid2(
    cols = vars(group),
    scales = "free_y",
    labeller = labeller(group = function(x) {
      recode(x, `1` = "Currently virgin", `2` = "Current sexual debuted", `3` = "Currently married")
    })
  ) +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, .1)) +
  guides(size = FALSE) +
  labs(x = 'Age', y = 'Transition probability', color = 'Transition') +
  k.theme + k.legend(.925, .5, dir = "vertical")
```

::: {#ref}
:::

# Appendix

```{r}
#| label: fig-exampleHazard
#| fig-cap: Illustration of the Log-logistic base hazard with exponential decay (EDLL) hazard function for $\lambda = \exp(8)$, $\beta = 2.5$, $\delta = 0.01$, and $k = 0.1875$. The hazard is defined as $h(t; \lambda, \beta, \delta, k) = \frac{\beta t^{\beta-1}}{(\lambda + t^\beta)(1+\delta e^{kt})}$. 
#| include: true
#| fig-env: "figure*"
#| fig-pos: "!ht"
#| fig-width: 7
#| fig-height: 5
#| out-width: "80%"

lambda <- exp(8)
beta   <- 2.5
delta  <- 0.01
k_val  <- 0.1875 

# Log-hazard
# h(t) = (beta * t^(beta-1)) / ((lambda + t^beta) * (1 + delta * exp(k*t)))
log_h_t_calculator <- function(t, lambda, beta, delta, k) {
  logspace_add <- function(logx, logy) 
    pmax(logx, logy) + log1p(exp(-abs(logx - logy)))
  log_denom1 <- logspace_add(log(lambda), beta * log(t))
  log_denom2 <- logspace_add(0, log(delta) + k * t)
  log(beta) + (beta - 1) * log(t) - log_denom1 - log_denom2
}

log_h_t_calculator(1:50, lambda, beta, delta, k_val) %>%
  exp() %>%
  plot(1:50, .,
    type = "l",
    xlab = "Time (t)", ylab = "Hazard h(t)",
    main = "Simulated Hazard Function"
  )
```
