--- 
title: "Rate of sexual debut, marriage, and marital dissolutions in Sub-Saharan Africa"
author:
  - name: Văn Kính Nguyễn
    id: vkn
    email: kinh.nguyen@uni-heidelberg.de
    affiliation:
      - ref: 1
  - name: Jeffrey Imai-Eaton
    affiliation:
      - ref: 2
affiliations: 
  - id: 1
    department: Heidelberg Institute for Global Health
    name: Heidelberg University Hospital
    city: Heidelberg
    state: Baden-Würtenburg
    country: Germanty
    address: Im Neuenheimer Feld 130/3
    postcode: 69120
    url: uni-heidelberg.de
  - id: 2
    department: Imperial College London
    name: Heidelberg University Hospital
    city: Heidelberg
    state: Baden-Würtenburg
    country: Germanty
    address: Im Neuenheimer Feld 130/3
    postcode: 69120
    url: uni-heidelberg.de

execute: 
  cache: true
  warning: false
  error: false
  include: false
  echo: false

editor:
  render-on-save: true

format:
  elsevier-pdf:
    journal:
      name: "medArxiv"
      formatting: preprint
      model: 3p
      layout: twocolumn
    number-sections: false
    include-before-body:
      text: |
        \linenumbers
    include-in-header:
      text: |
        \usepackage{amsmath}
        %\usepackage[affil-it]{authblk}
        %\renewcommand{\Affilfont}{\footnotesize}
        \usepackage{hyperref}
        \hypersetup{colorlinks, linkcolor={blue}, citecolor={blue}, urlcolor={blue}}
        \usepackage[capitalise,noabbrev]{cleveref}
        \usepackage[title]{appendix}
        \usepackage[inline]{enumitem}
        %\usepackage{paralist}
        \usepackage{blkarray} % named matrixed
        \usepackage{lineno}
        \usepackage{etoolbox}
        \AtBeginEnvironment{abstract}{\setlength{\parindent}{0pt}}
        %\usepackage[hpos=0.15\paperwidth,vpos=0.1\paperheight,colormodel=RGB,color={[RGB]{253, 246, 227}}]{draftwatermark}
        %\SetWatermarkText{draft}
        %\SetWatermarkScale{1}
        %\usepackage{xcolor}
        \definecolor{pageDark}{rgb}{0.992,0.9647,0.8902}
        \definecolor{textDark}{rgb}{0.396,0.482,0.514}
        \definecolor{pageLight}{rgb}{0.0,0.1686,0.2117}
        \definecolor{textLight}{rgb}{0.5137,0.5804,0.5882}
        \pagecolor{pageDark}
        \color{textDark} 
        %\pagecolor{pageLight}
        %\color{textLight} 
    # linestretch: 1
    pagestyle: headings
    fontsize: 11pt
    colorlinks: true
    indent: false
    keep-tex: true

bibliography: "/Users/knguyen/Documents/zotero.bib"
latex-clean: false
fig-dpi: 300
abstract: > 
  **Background** 
  This paper details a continuous-time multistate model designed to analyze
  transitions through various life course stages, including sexual debut,
  marriage, marital dissolution (separation, divorce, widowhood), and
  remarriage. The model incorporates age-dependent transition intensities and
  allows for the inclusion of other covariates. We discuss the theoretical
  underpinnings, model specification, calculation of transition matrices, and
  the likelihood framework, particularly focusing on an approach suitable for
  panel data where exact transition times for all states may not be observed.
---

# Background

Tireless efforts in HIV/AIDS research and treatment have led to better life for
people living with HIV (PLHIV). Treatment is yet still a lifelong process, and
every year, new infections, while have been subsided since 1996, still amount to
more than a million cases globally [@unaids2023]. The new infections are
disproportional distributed among women of young age [@unaids2023], the period
during which multiple sexual related life events occur including sexual debut,
marriage, and marriage dissolution.

In the marital status categories, the HIV prevalence has been shown to be
higher. Remarried individuals form a large portion of the population and have a
higher-than-average HIV prevalence [@de_walque_association_2012], whereas widows
and widowers had exceptionally high HIV prevalence [@mcgrath_older_2009]. The
risk of acquiring HIV increased in remarriage couples [@maphosa_unmasking_2014]
and was significantly associated with being married or in one of the marriage
dissolution states [@kimani_marital_2013]. 

Social and cultural reasons could lead to the increasing risk and prevalence in
those groups. HIV-positive surviving spouses might not want to disclose their
status and lessen the precautionary measures as children could be expected from
the new marriages. Widowhood could lead to high-risk partnerships to be formed
when widowed men and women reconnect to the sexual network
[@adetunji_hivaids_2001;@lopman_role_2009]. A higher rate of partner acquisition
could be expected due to polygamous marriage, frequency of remarriage and
marriage dissolution. Those with multiple marriages have a higher probability of
HIV infection among women with circumcision spouses [@poulin_inquiry_2011]. A
widespread cultural practice in sub-Saharan Africa that could also contribute to
risk of HIV transmission is widow inheritance [@agot_widow_2010].

The population of HIV-positive partners, surviving spouses, and divorcee could
face more difficulties to to access supporting services. In many African
countries, divorce is seen as a sign of failure and social ostracism often
follows which can lead to marginalization. Often lack of education and have
limited incomes, divorced and widowed women could be more likely to engage in
transactional sex or multiple concurrent partnerships and unable to negotiate
safe sex practices with partners. They might also not have access to prevention
services because they do not have the legal rights required for accessing these
services without male support. Unaddressed, these individuals would not only
face a higher risk of life-threatening conditions but also fuels the continuing
of the epidemic. 

Joint actions by governments and international organizations are needed to
better understand the gap and address the needs of different marital status
populations. In order to understand the extends of the resources needed as well
as potential impacts of programs targeting the population, estimating of the
life-course transitions between the marital statuses are needed. Improved
estimates of the transition rates will inform the estimate of the target
population size and dynamic models for epidemic modelling and assessments.

# Materials & Methods

We used a multistate survival framework to model the rates of transition between
marital statuses. Multistate models (MSMs) is a robust framework for analyzing
the movement through a series of discrete states over time by specifying the
instantaneous risk of moving from state $i$ to state $j$ at time $t$. To
characterize temporal trends and geographic variation in the rate, we allowed
the parameters to vary by country and survey years. Changes over time were
represented by a second-order auto-regressive (AR2). The model was fitted
separately for male and female.

## Household survey data

Data on age at first sex (AFS), age at marriage (AAM), and current marital
status were obtained from Demographic and Health Surveys (DHS) in 37 countries.
Only nationally representative surveys with individual records available for
analysis were used. For AFS, respondents stated their age at first sexual
intercourse; valid responses were 'never,' an integer age, or 'at first
marriage/union,' which was substituted with the reported age at first marriage.
Respondents also recalled their age at first marriage or union. Recorded current
marital statuses were: "Never married nor lived in a consensual union," "Legally
or formally married," "Not legally or formally married but living with a
man/woman in a consensual union," "Widowed from a marriage or consensual union
and not remarried or not in a consensual union," "Divorced from a legal or
formal marriage and not remarried or in a consensual union," "Separated from a
marriage or consensual union," and "Married or living in a union."

## States transition

We defined seven states of transition from virgin from birth ($V$) to current
marital status sexually debuted (and not married) ($X$), married or in a union
($M$), three states of marriage dissolution including separated ($S$), divorced
($D$), widowed ($W$), and remarried or in a union again ($R$). A remarried event
is defined if the respondent reported that there had been more than one union
and currently is married, in a union, or is in one of the three stages of
marriage dissolution. 

Transitions are unidirectional for initial life stages and allow for cycles in
later marital statuses: V $\to$ X, V $\to$ M (direct marriage implies debut), X
$\to$ M, M $\to$ S, M $\to$ D, M $\to$ W, S $\to$ D, S $\to$ R, D $\to$ R, W
$\to$ R, R $\to$ S, R $\to$ D, R $\to$ W. The states S, D, W, and R can form
cycles. Furthermore, when the AFS is reported equal or greater than AAM we set
the transition as the respondent moved from a virgin directly to married. There
can be as many number of unions as theoretically allowed to traverse the paths
in those records with more than once unions. However, while marital dissolution
states are not observed, we assumed if a marriage dissolute within a year after
married, there was no intermediate states in between, except the current status 
is remarried.

## Transition Intensities and Probabilities

The instantaneous transition intensity from state $i$ to state $j$ at age $t$,
given a vector of covariates $\mathbf{X}$, is modeled using a linear predictor
on the log scale, equivalent to a Gompertz hazards type model [@van2016multi],
i.e., 
\begin{equation}\label{eq:eta}
\lambda_{ij}(t, \mathbf{X}) = \alpha_{ij} + \beta_{ij} t + \sum_k \gamma_{ijk}
X_k
\end{equation}
where $\alpha_{ij}$ is the baseline intensity intercept, $\beta_{ij}$ is
the age effect, and $\gamma_{ijk}$ are coefficients for other time-constant or
time-varying covariates $X_k$. The parameters $\lambda_{ij}$ are the
instantaneous transition rates between the states, which is described with the
intensity matrix $Q(t, \mathbf{X})$, a square matrix where
\begin{equation}\label{mat:Q}
\scriptsize
Q = 
\begin{blockarray}{cccccccc}
V & X & M & S & D & W & R \\
\begin{block}{(ccccccc)c}
  q_{11} & \lambda_{12} & \lambda_{13} & 0 & 0 & 0 & 0 & V \\
  0 & q_{22} & \lambda_{23} & 0 & 0 & 0 & 0 & X \\
  0 & 0 & q_{33} & \lambda_{34} & \lambda_{35} & \lambda_{36} & 0 & M \\
  0 & 0 & 0 & q_{44} & 0 & 0 & \lambda_{47} & S \\
  0 & 0 & 0 & 0 & q_{55} & 0 & \lambda_{57} & D \\
  0 & 0 & 0 & 0 & 0 & q_{66} & \lambda_{67} & W \\
  0 & 0 & 0 & \lambda_{74} & \lambda_{75} & \lambda_{76} & q_{77} & R \\
\end{block}
\end{blockarray}
\end{equation}
The off-diagonal elements $q_{ij}(t, \mathbf{X}) = \lambda_{ij}(t, \mathbf{X})$
for $i \neq j$, representing the intensity of transition from state $i$ to state
$j$. The diagonal elements $q_{ii}(t, \mathbf{X}) = - \sum_{j \neq i}
\lambda_{ij}(t, \mathbf{X})$, representing the negative of the total rate of
leaving state $i$, e.g., 
$q_{11}(t) = -(\lambda_{V,SD}(t) + \lambda_{V,M1}(t))$.

The remarried state $R$ is separated from $M$ to model individuals who reported
having more than once union. While the rate of marital dissolution
could be different from those who had married only once ($M$) and those had
remarried ($R$), all the marriage dissolution events were not observed in the
data. Thus, we set the marital dissolution rates to be the same for both
marriage and remarried states $M$ and $R$, i.e., $q_{33} = q_{77},\ \lambda_{34}
= \lambda_{74},\ \lambda_{35} = \lambda_{75},\ \lambda_{36} = \lambda_{76}$. In
addition, the three remarried rates from either of the three marriage
dissolution states ($S, D, W$) are only informed by the number of individuals
who reported having more than once union and currently married --- it is unknown
from which state the individual had remarried. Thus, we constrained the three
rates to be equal of the parameters, i.e., $\lambda_{47} = \lambda_{57} =
\lambda_{67}$.

## Model likelihood

The likelihood is based on individuals' observed histories which includes
segments of history where the state occupancy is continuously known and segments
where only the start and end states of an interval are known. 
Let $t_0$ be the initial age of observation, $t_c$ be the current age at
observation, $t_d$ be the age at sexual debut, and $t_{m}$ be the age at
first marriage. 
The survival probability in state $i$ from $t_a$ to $t_b$ is $\exp(-H_i(t_a,
t_b, \mathbf{X}))$ [@van2016multi], where 
\begin{flalign*}
H_i(t_a, t_b, \mathbf{X}) &= \int_{t_a}^{t_b} q_{ii}(u, \mathbf{X}) du
%& = \int_{t_a}^{t_b} -\sum_{j \neq i} \lambda_{ij}(u, \mathbf{X}) du
%&= -\left[ A_i(\mathbf{X})(t_b-t_a) + B_i \frac{t_b^2-t_a^2}{2} \right] \\
\end{flalign*}
For currently virgin observations, the individual has remained in state V from
$t_0$ to $t_c$  with
$$L_V = \exp(-H_V(t_0, t_c, \mathbf{X})).$$
Those made sexual debut at $t_d$, the path is then V($t_0$) $\to$ V($t_d^-$)
$\to$ X($t_d$) $\to$ X($t_c$) with the likelihood contribution as
\begin{align*}
L_{X} = & \exp(-H_V(t_0, t_d, \mathbf{X})) \times \lambda_{V,X}(t_d, \mathbf{X}) \\
        \times & \exp(-H_{X}(t_d, t_c, \mathbf{X})).
\end{align*}
Those are currently in their first marriage at $t_c$, there are two
sub-paths: who made sexual debut before married with transition V $\to$ X $\to$
M ($t_d < t_{m}$) has 
\begin{align*}
L_{M}^{(1)} & = \exp(-H_V(t_0, t_d, \mathbf{X})) \times \lambda_{V,X}(t_d,
\mathbf{X}) \\
& \times \exp(-H_{X}(t_d, t_{m}, \mathbf{X})) \times \lambda_{X,M}(t_{m}, \mathbf{X}) \\
& \times \exp(-H_{M}(t_{m}, t_c, \mathbf{X})),
\end{align*}
whereas those have $t_d = t_{m}$ or unknown $t_d$ are assumed transits V $\to$
M has
\begin{align*}
L_{M}^{(2)} & = \exp(-H_V(t_0, t_{m}, \mathbf{X})) \times \lambda_{V,M}(t_{M}, \mathbf{X}) \\
  & \times \exp(-H_{M}(t_{m}, t_c, \mathbf{X})).
\end{align*}
For those who are currently separated (S), divorced (D), widowed (W), or
remarried (R) at $t_c$, there are two phases: likelihood of reaching M at
$t_{m}$ either $L_{M1}^{(1)}$ or $L_{M1}^{(2)}$ 
\begin{align*}
L_{M}^{(1)} = & \exp(-H_V(t_0, t_d, \mathbf{X})) \lambda_{V,X}(t_d, \mathbf{X})  \\
& \exp(-H_{X}(t_d, t_{m}, \mathbf{X})) \lambda_{X,M}(t_{m}, \mathbf{X}).\\
L_{M}^{(2)} = & \exp(-H_V(t_0, t_m, \mathbf{X})) \lambda_{V,M}(t_m, \mathbf{X}).
\end{align*}
and likelihood of being in state $C \in \{S, D, W, R\}$ at $t_c$, given in M
at $t_{m}$. 
$$L_C = L_{M}^{(*)} \times (P(t_{m}, t_c, \mathbf{X}))_{M, C}$$
where $P(\cdot)_{i,j}$ an element from the probability transition matrix P-matrix.
The term $P$ accounts for all possible sequences of transitions that could occur
between $t_m$ and $t_{c}$ starting in state M and ending in state C
[@van2016multi;@jackson2022].

The P-matrix contains elements $p_{ij}(t_1, t_2, \mathbf{X})$ representing the
probability that an individual in state $i$ at time $t_1$ will be in state $j$
at time $t_2$. The matrix is computed by solving the differential equations:
$$
\frac{d}{dt} P(t_1, t, \mathbf{X}) = P(t_1, t, \mathbf{X}) Q(t, \mathbf{X})
$$
with the initial condition $P(t_1, t_1, \mathbf{X}) = \mathbf{I}$ 
to obtain $P(t_1, t_2, \mathbf{X})$.
When $Q(t, \mathbf{X})$ is
constant over the interval $[t_1, t_2]$, then $P(t_1, t_2, \mathbf{X}) =
\text{expm}(Q \cdot (t_2 - t_1))$, where
`expm` is the matrix exponential [@van2016multi].
Here, since $Q(t, \mathbf{X})$ is age-dependent, $P(t_1, t_2, \mathbf{X})$ is
computed by dividing the interval $[t_1, t_2]$ into small subintervals $\Delta
t_k$ within which $Q(t_k, \mathbf{X})$ is assumed constant. In particular,
$$ P(t_1, t_2, \mathbf{X}) \approx \prod_{k} \text{expm}(Q(t_k, \mathbf{X})
\cdot \Delta t_k)$$
Here, we use $\Delta t_k = 1$ year, and $Q(t_k, \mathbf{X})$ is
evaluated at the start of the year $t_k$.

## Computational

Calculating the matrix exponential is an expensive operation
[@molerNineteenDubiousWays2003], thus, for each parameter set, we pre-computed
the values for all the unique combinations of individual profile and only
queried the values during the likelihood calculation. Since there were over a
million individual records, to reduce the size of the data for computational
purposes, the standard time-split data format (each row an episode) was not used
but indicators for the likelihood's terms in \cref{eq_likelihood} were
calculated for each individual (each row an individual). Furthermore,
individuals with similar life-course were further aggregated to group in which
the group counts was used to weight the likelihood in \cref{eq_likelihood}, thus
reducing the number of unique calculation of the likelihood.

The model and likelihood calculation were implemented with Template Model
Builder which can be accessed on Github's repository
[kinh-nguyen/multistage]{https://github.com/kinh-nguyen/multistage}. Parameter
estimate was done with empirical Bayes. The priors were chosen to shrink the
regression coefficients towards zero when they were only weakly informed by the
data. In particular, $\beta_{0,rs} \sim N(0, 0.5)$, $\beta_{1, rs}\sim N(0,
0.1)$ and $c_(\text{country})\sim N(0, 0.1)$.  We computed posterior medians and
95\% credible intervals (95\% CI) of the outcomes of interests, including the
rate, transition probability, and hazard by ages, using posterior samples of
3000 parameters.

# Results

```{r}
library(ktools)
library("TMB")
vroom::vroom(here("data/new_d.csv"), col_select = -1) %>% 
  mutate(
    A = case_when(A == 'V' ~ 0, A == "X" ~ 1, A == "M" ~ 2, A == 'S' ~ 3, A == "D" ~ 4, A == "W" ~ 5, A == "R" ~ 6),
    Z = case_when(Z == 'V' ~ 0, Z == "X" ~ 1, Z == "M" ~ 2, Z == 'S' ~ 3, Z == "D" ~ 4, Z == "W" ~ 5, Z == "R" ~ 6),
    across(where(is.numeric), as.integer)
    ) %>% 
allot(d)
```

```{r, modelfit}
#| eval: false
# recode unobserved stages to repeated calculation
d %>% 
  mutate(
    fit = case_when(
      # Q(t,X)[A,Z]: observed transition at start time, within a year, except
      # remarried is current state
      A != Z & (end - start == 1) & Z != 6 ~ 0,
      # P(t,X)[A,Z], A == Z, remain the same
      A == Z & (end - start == 1) ~ 1,
      # (P(start,X) x P(start+1,X) x ... x P(end, X))[A,Z]: unobserved
      # transition, matrix multiplication
      # A != Z & ((end - start > 1) | Z == 6) ~ 2,
      otherwise ~ 2
    )
  )  %>% 
allot(d1)
N_PAR = 7
fit_ct = "MW"
fitdt <- filter(d1, sex == 'female', cc == fit_ct) %>% 
  filter(cc == fit_ct, sex == 'female') %>% 
  group_by(cc, sex, A, Z, start, end, fit) %>%
  summarise(n = sum(n)) %>% # remove surveys effects
  ungroup()

tdt <- fitdt %>%
  select(A, Z, start, end, n, fit)
```

```{r, modelfit}
data <- list()
data$prior_base <- c(0, 2.5) # log normal mean and sd
data$prior_t <- c(0, 2.5) # mean and sd

data$n_age <- max(tdt$end) + 1 # include zero

# compute the length of age_sm
tdt %>%
  mutate(maxage = max(end)) %>% # although there is no e.g. 50-51, but can be 20-51 jump
  group_by(A, Z) %>%
  reframe(
    minage = min(start),
    maxage = max(maxage),
    nepis = maxage - minage + 1
  ) %>%
  filter(A != Z) %>%
  arrange(A, Z) %>%
allot(exist_age)

age_dv = rep(0, N_PAR * data$n_age)
age_pv = rep(1, sum(exist_age$nepis))
data$age_dv = age_dv
data$minage = exist_age$minage
data$n_epis = exist_age$nepis
data <- modifyList(data, as.list(tdt))

init <- list(
    intercepts = rnorm(N_PAR, data$prior_base[1], data$prior_base[2]), 
    log_sigma_rw1 = log(1), 
    age_sm = age_pv
)
library(TMB)
TMB::compile("code/msm.cpp", flags = "-Wno-ignored-attributes")
base::dyn.load(TMB::dynlib("code/msm"))
invisible(TMB::config(tape.parallel = 0, DLL = "msm"))
obj <- TMB::MakeADFun(data, init, DLL = "msm", silent = TRUE)
sml <- obj$simulate() 
fit <- nlminb(obj$par, obj$fn, obj$gr, control = list(trace = 1))
rp <- obj$simulate(obj$env$last.par.best)
dir.create(here('fit'), showWarnings = FALSE)
saveRDS(rp, here('fit/testMW.rds'))
```

```{r, loadResult}
rp <- readRDS(here('fit/testMW.rds'))
P_yearly = rp$pM_rep
pe = tibble(
  start = 0:51,
  VV = P_yearly[1,1,],
  VX = P_yearly[1,2,],
  VM = P_yearly[1,3,],
  XX = P_yearly[2,2,],
  XM = P_yearly[2,3,],
  MM = P_yearly[3,3,],
  MS = P_yearly[3,4,],
  MD = P_yearly[3,5,],
  MW = P_yearly[3,6,],
  MR = P_yearly[3,7,],
) %>% 
pivot_longer(-start)
```

```{r}
#| label: fig-fittedExample
#| fig-cap: Fitted example
#| include: true
#| fig-env: "figure*"
#| fig-pos: "!ht"
#| fig-width: 7
#| fig-height: 6
#| out-width: "100%"

d %>% 
  filter(cc == "ZA", sex == 'female') %>% 
  mutate(duration = end - start) %>% 
  group_by(A, start, duration) %>%
  summarise(N_A_start_duration = sum(n), .groups = 'drop') %>% 
allot(at_risk)

d %>% 
  filter(cc == "ZA", sex == 'female') %>% 
  mutate(duration = end - start) %>% 
  group_by(A, start, duration) %>%
  summarise(N_A_start_duration = sum(n), .groups = 'drop') %>% 
allot(at_risk)

d %>% 
  filter(cc == "ZA", sex == 'female') %>% 
  mutate(duration = end - start) %>% 
  left_join(at_risk, by = c("A", "start", "duration")) %>%
  mutate(p_empirical = n / N_A_start_duration) %>% 
  mutate(
    A = case_when(A == 0 ~ 'V', A == 1 ~ 'X', A == 2 ~ "M", A == 3 ~ "S", A == 4 ~ "D", A == 5 ~ "W", A == 6 ~ 'R', otherwise ~ NA_character_),
    Z = case_when(Z == 0 ~ 'V', Z == 1 ~ 'X', Z == 2 ~ "M", Z == 3 ~ "S", Z == 4 ~ "D", Z == 5 ~ "W", Z == 6 ~ 'R', otherwise ~ NA_character_),
    name = paste0(A,Z)) %>% 
ggplot(aes(start, p_empirical, color = factor(name))) +
  geom_point() +
  geom_line(data = pe, aes(y = value)) +
  facet_wrap(~name, scales = 'free_y') +
  scale_y_continuous(labels = scales::percent) +
  geom_hline(yintercept = 0) +
  guides(color = FALSE)
```

::: {#ref}
:::
