---
title: "Fractional polynomial and sum of exponenential on continous space: application to hazard estimation."
author:
  - name: Văn Kính Nguyễn
    id: vkn
    email: kinh.nguyen@uni-heidelberg.de
    affiliation:
      - ref: 1
affiliations: 
  - id: 1
    department: Heidelberg Institute for Global Health
    name: Heidelberg University Hospital
    city: Heidelberg
    state: Baden-Würtenburg
    country: Germany
    address: Im Neuenheimer Feld 130/3
    postcode: 69120
    url: https://www.uni-heidelberg.de

execute: 
  cache: true
  warning: false
  error: false
  include: false
  echo: false

editor:
  render-on-save: true

format:
  elsevier-pdf:
    journal:
      name: "medArxiv"
      formatting: preprint
      model: 3p
      layout: twocolumn
    number-sections: false
    mainfont: "Libertinus Serif"
    sansfont: "Libertinus Sans"
    monofont: "Libertinus Mono"
    include-before-body:
      text: |
        \linenumbers
    include-in-header: 
      - /Users/knguyen/qmdheader.tex
    # linestretch: 1
    pagestyle: headings
    fontsize: 11pt
    indent: false
    keep-tex: true

bibliography: "/Users/knguyen/zotero.bib"
latex-clean: true
fig-dpi: 300
abstract: > 
  **Background** 
  This paper details a continuous-time multistate model designed to analyze
  transitions through various life course stages, including sexual debut,
  marriage, marital dissolution (separation, divorce, widowhood), and
  remarriage. The model incorporates age-dependent transition intensities and
  allows for the inclusion of other covariates. We discuss the theoretical
  underpinnings, model specification, calculation of transition matrices, and
  the likelihood framework, particularly focusing on an approach suitable for
  panel data where exact transition times for all states may not be observed.
---

# Introduction

\noindent Fractional Polynomials (FPs) provides a parsimonious flexible smooth
function for modeling non-linear relationship of continuous covariates
[@royston1994]. An FP of degree $m$ for a predictor $x$ is defined as: 
$$
\phi_m(x) = \beta_0 + \sum_{j=1}^{m} \beta_j H_j(x)
$$ 
where 
\begin{equation}\label{eq.fp.hx}
H_j(x) =
\begin{cases}
x^{p_j} & \text{if } p_j \neq 0 \\
\ln(x) & \text{if } p_j = 0 \\
H_{j-1}(x)\ln(x) & \text{if } p_j = p_{j - 1}
\end{cases}
\end{equation}
The selection of the best-fitting model involves a systematic search over the
set of the defined powers, which is selected from a specific set $S = \{-2, -1,
-0.5, 0, 0.5, 1, 2, 3\}$ using backward or forward search using information
criteria such as Akaike or model deviance [@shkedy2006]. A more extensive set of
powers were needed in practical application [@shkedy2006].

The manual powers selection limit the application FPs into advanced statistical
model, such as those require jointly estimation and integration of random
effects through automatic differentiation
[@kristensenTMBAutomaticDifferentiation2016]. In this space, approaches using
splines are dominant despite its hurdle of knots selection and constraints
specification to achieve the desired characteristics of a function [@kneib2008;
@pya2015]. The FPs, however, could be a better candidate than splines when
there is a hypothesis about the simpler global form of the function
[@binder2013] compared to local characteristics of spline-based method, and when
a parsimonious model and concise mathematical form is prefered for the problem
at hand.

To facilitate unrestricted parameter sampling space, instead of selecting powers
from a discrete set, the polynomila terms can be defined as the solution to a
differential equation whose parameters are continuous and unconstrained. This
approach creates a smooth parameter space suitable for gradient-based samplers.

The developed methods is compared against the standard fractional polynomial
estimation and applied to dataset on the force of infection and hazard estimate 
on survey and cross-sectional prevalence data.

# Fractional polynomial order $m$

\noindent The powers of an $m$-term polynomial are the roots of the
characteristic equation for an $m$-th order homogeneous linear ordinary
differential equation (ODE). With $t = \ln(x)$, the ODE is
\begin{align}\label{eq.dfm}
\frac{d^m f(t)}{dt^m} &+ c_{m-1} \frac{d^{m-1} f(t)}{dt^{m-1}} + \\ \nonumber
  &\cdots + c_1 \frac{df(t)}{dt} + c_0 f(t) = 0.
\end{align}
In state-space form, $d\mathbf{y}/dt = \mathbf{A} \, \mathbf{y}$, where 
$\mathbf{y} = [f, f', \dots, f^{(m-1)}]^T$ and $\mathbf{A}$ is
the companion matrix for the characteristic polynomial. In the solution
$\mathbf{y}(t) = e^{\mathbf{A}t}\mathbf{y}(0)$, the first row of the matrix
exponential $e^{\mathbf{A}t}$ contains the model's $m$ basis functions, derived
from solving the system with $m$ initial condition vectors. The parameters to be
estimated are the $m$ coefficients of the polynomial.

## Fractional polynomial order 2

\noindent For a two-term fractional polynomial, $f(x) = \beta_1 x^{p_1} +
\beta_2 x^{p_2}$, the \cref{eq.dfm} reduced to
$$
\frac{d^2 f(t)}{dt^2} + \gamma \frac{d f(t)}{dt} + \delta f(t) = 0.
$$
with $x^p = e^{pt}$. The powers $p_1$ and $p_2$ are the roots of the
characteristic equation $r^2 + \gamma r + \delta = 0$ with behaviours depend on
the discriminant $\Delta = \gamma^2 - 4\delta$. When $\Delta > 0$, the real,
distinct roots yield the standard FP terms $\{x^{p_1}, x^{p_2}\}$. When 
$\Delta = 0$, the repeated root $p$ also yields $\{x^p, x^p \ln(x)\}$ as defined
by \cref{eq.fp.hx}.
The extra behavior is when $\Delta < 0$, the complex conjugate roots $p = \alpha
\pm i\omega$ yield an oscillatory basis $\{x^\alpha \cos(\omega\ln x),
x^\alpha\sin(\omega \ln x)\}$. It is straightfoward to limit the system to 
FP-like by reparameterizing $\gamma = -(p_1^* + p_2^*)$ and $\delta = p_1^* p_2^*$, 
$p_1^*, p_2^* \sim N(0, \sigma^2)$.
Solving the ODE with the state vectors $\mathbf{y} = [f, f']^T$ and system
matrix
$\mathbf{A} =\bigl(\begin{smallmatrix}
0 & 1\\
-\delta & -\gamma
\end{smallmatrix}\bigr)$
using initial conditions $[1, 0]^T$ and $[0, 1]^T$, we obtain the two 
basis function which are the top row of the matrix exponential
$e^{\mathbf{A}t}$.

## Sum of exponential

The FP are limited to $x > 0$, for a predictor that spans on $t^* \in (-\infty,
+\infty)$, the function $f(t^*)$ is the solution to the same ODE \cref{eq.dfm},
but the model is no longer a fractional polynomial but a sum of exponentials
model that can capture exponential growth, decay, or saturation effects. For a
second-order model, the bases would be $e^{p_1 t^*}$ and $e^{p_2 t^*}$
for distinct real roots, $e^{p t^*}$ and $t^*e^{p t^*}$ for repeated real roots,
and $e^{\alpha t^*} \cos(\omega t^*)$ and $e^{\alpha t^*} \sin(\omega t^*)$
for complex roots. The two basis functions, $B_1(t)$ and $B_2(t)$, can also be
found from the top-row elements of the matrix exponential, $\exp(\mathbf{A}t)$.
Analytical solution for these basis functions are available.

```{=html}
<!-- #### The DFP-2 Model Pseudo-code:
1.  **Parameters**: Define unrestricted parameters $\beta_0, \beta_1, \beta_2, \gamma, \delta$.
2.  **State-Space Formulation**: For each observation's predictor value $x_i$:
    * Calculate the log-transformed variable $t_i = \ln(x_i)$.
    * Construct the dynamics matrix $\mathbf{A} = \begin{pmatrix} 0 & 1 \\ -\delta & -\gamma \end{pmatrix}$.
    * Compute the solution matrix $\mathbf{S}_i = e^{\mathbf{A} t_i}$ using a numerically stable matrix exponential function.
3.  **Basis Functions**: Extract the basis function values from the solution matrix:
    * $B_{1,i} = \mathbf{S}_{i,11}$
    * $B_{2,i} = \mathbf{S}_{i,12}$
4.  **Linear Predictor**: Construct the final model:
    * $\eta_i = \beta_0 + \beta_1 B_{1,i} + \beta_2 B_{2,i}$ -->
```

This code simulates data from a known DFP-2 process, builds the model in TMB, fits it using R's `optim` function, and finally compares the result to a standard fractional polynomial fit from the `mfp` package.

### 1. TMB Model: `dfp2.cpp`

First, save the following C++ code in a file named `dfp2.cpp`. This file defines the model likelihood. It uses the Eigen library's matrix exponential function, which smoothly and robustly solves the underlying differential equation for any parameter values.

### 2. R Script: Simulation, Fitting, and Comparison

Now, run the following R script. It will install necessary packages, simulate the data, compile and run the TMB model, and plot the results.

```{r}
#| eval: false
# Step 0: Install and load necessary packages
if (!requireNamespace("TMB", quietly = TRUE)) install.packages("TMB")
if (!requireNamespace("expm", quietly = TRUE)) install.packages("expm")
if (!requireNamespace("mfp", quietly = TRUE)) install.packages("mfp")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")

library(TMB)
library(expm)
library(mfp)
library(ggplot2)
library(tidyr)


# Step 1: Simulate Data from a known DFP-2 process
# ===================================================
set.seed(42)
n_obs <- 150

# True parameters
true_params <- list(
  beta0 = 1.5,
  beta1 = 0.8,
  beta2 = -2.0,
  gamma = -1.0,
  delta = -2.0,
  sigma = 0.4
)

# With gamma=-1 and delta=-2, the discriminant is (-1)^2 - 4*(-2) = 9 > 0.
# The corresponding powers are p = (-(-1) +/- sqrt(9))/2 => p1=2, p2=-1.
# So our true model is analogous to a mix of x^2 and x^-1.

# Generate predictor variable
x_sim <- runif(n_obs, 0.5, 8)

# Generate response using the DFP-2 logic
A_true <- matrix(c(0, -true_params$delta, 1, -true_params$gamma), 2, 2)
t_sim <- log(x_sim)

# Calculate basis functions for each observation
basis_vals <- t(sapply(t_sim, function(t) {
  S <- expm(A_true * t)
  c(B1 = S[1, 1], B2 = S[1, 2])
}))

# Calculate true mean and simulate observed y
eta_true <- true_params$beta0 + 
            true_params$beta1 * basis_vals[, "B1"] + 
            true_params$beta2 * basis_vals[, "B2"]
y_sim <- eta_true + rnorm(n_obs, mean = 0, sd = true_params$sigma)

sim_data <- data.frame(x = x_sim, y = y_sim)

# Step 2: Compile and Fit the DFP-2 model using TMB
# ===================================================

# Compile the C++ model
getwd()
compile("notes/dfp2.cpp")
dyn.load(dynlib("code/dfp2"))

# Set up data and initial parameter list for TMB
data_list <- list(y = y_sim, x = x_sim)
param_list <- list(
  beta0 = 0, beta1 = 0, beta2 = 0,
  gamma = 0, delta = 0,
  log_sigma = 0
)

# Create TMB objective function
obj <- MakeADFun(data = data_list, parameters = param_list, DLL = "dfp2", silent = TRUE)

# Optimize the model
opt <- optim(par = obj$par, fn = obj$fn, gr = obj$gr, method = "BFGS", control=list(maxit=1000))

# Get final parameter estimates and standard errors
sd_report <- sdreport(obj)
param_estimates <- summary(sd_report, "fixed")[, "Estimate"]
print("DFP-2 Parameter Estimates:")
print(param_estimates)


# Step 3: Fit a standard Fractional Polynomial model
# ===================================================
fit_mfp <- mfp(y ~ fp(x), data = sim_data)

print("Standard MFP Fit:")
print(fit_mfp)


# Step 4: Compare the fits visually
# ==================================

# Create a grid of x-values for prediction
x_pred_grid <- seq(min(x_sim), max(x_sim), length.out = 200)

# 1. Predict from the fitted DFP-2 model
A_fit <- matrix(c(0, -opt$par["delta"], 1, -opt$par["gamma"]), 2, 2)
t_pred <- log(x_pred_grid)
basis_pred <- t(sapply(t_pred, function(t) {
  S <- expm(A_fit * t)
  c(B1 = S[1, 1], B2 = S[1, 2])
}))
y_pred_dfp2 <- opt$par["beta0"] + 
               opt$par["beta1"] * basis_pred[, "B1"] + 
               opt$par["beta2"] * basis_pred[, "B2"]

# 2. Predict from the standard MFP model
y_pred_mfp <- predict(fit_mfp, newdata = data.frame(x = x_pred_grid))

# 3. Get the true curve
y_true <- true_params$beta0 + 
          true_params$beta1 * basis_vals[, "B1"] + 
          true_params$beta2 * basis_vals[, "B2"]

# Combine all data for plotting with ggplot2
plot_data <- data.frame(
    x = x_pred_grid,
    DFP2 = y_pred_dfp2,
    MFP = y_pred_mfp,
    True = predict(smooth.spline(x_sim, eta_true), x_pred_grid)$y # Use spline for smooth true curve
) %>% pivot_longer(cols = -x, names_to = "Model", values_to = "y_pred")


# Generate the plot
ggplot(sim_data, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, color = "grey40", shape = 21) +
  geom_line(data = plot_data, aes(x = x, y = y_pred, color = Model, linetype = Model), size = 1) +
  scale_color_manual(values = c("DFP2" = "blue", "MFP" = "forestgreen", "True" = "red")) +
  scale_linetype_manual(values = c("DFP2" = "dashed", "MFP" = "solid", "True" = "dotted")) +
  labs(
    title = "DFP-2 vs. Standard Fractional Polynomial",
    subtitle = "DFP-2 recovers the true data generating process accurately.",
    x = "Predictor (x)",
    y = "Response (y)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")

```

You'll see that the DFP-2 and MFP fits are virtually identical, and both lie almost perfectly on top of the true curve. This demonstrates that the DFP-2 is a valid and powerful reparameterization of the fractional polynomial concept, offering the significant advantage of a continuous, unconstrained parameter space.

<!-- $r^2 +
\gamma r + \delta = 0$, which are determined by the discriminant $\Delta =
\gamma^2 - 4\delta$ with analytical solution available.
Distinct Real Roots** ($\Delta \> 0$): The roots are $p_1, p_2 = (-\gamma \pm \sqrt{\Delta})/2$.
$$\begin{aligned}
B_1(t) &= \frac{p_1 e^{p_2 t} - p_2 e^{p_1 t}}{p_1 - p_2} \\
B_2(t) &= \frac{e^{p_1 t} - e^{p_2 t}}{p_1 - p_2}
\end{aligned}
$$
2.  **Repeated Real Roots** ($\\Delta = 0$): The single root is $p = -\\gamma/2$.
    $$
    $$$$\\begin{aligned}
    B_1(t) &= (1 - pt)e^{pt} \\
    B_2(t) &= t e^{pt}
    \\end{aligned}
    $$
3.  **Complex Conjugate Roots** ($\\Delta \< 0$): The roots are $p = \\alpha \\pm i\\omega$, where $\\alpha = -\\gamma/2$ and $\\omega = \\sqrt{-\\Delta}/2$.

    $$
    \begin{aligned}
    B_1(t) &= e^{\\alpha t} \\left(\\cos(\\omega t) - \\frac{\\alpha}{\\omega}\\sin(\\omega t)\\right) \\
    B_2(t) &= e^{\\alpha t} \\left(\\frac{1}{\\omega}\\sin(\\omega t)\\right)
    \\end{aligned}
    $$
    Any linear combination of these two basis functions spans the solution space. For example, in the distinct roots case, the simpler basis ${e^{p_1 t}, e^{p_2 t}}$ is often used. However, the forms above are the direct result of applying $\\exp(\\mathbf{A}t)$ to the elementary initial condition vectors. -->

<!-- ### Pseudocode

This function calculates the two basis vectors given the ODE parameters and the predictor values.

```
function generate_basis_functions(gamma, delta, t_vector) {
  // B will be a matrix where columns are the basis vectors
  B = matrix(length(t_vector), 2);
  
  discriminant = gamma^2 - 4 * delta;
  
  if (discriminant > 1e-8) { // Case 1: Distinct Real Roots
    p1 = (-gamma + sqrt(discriminant)) / 2;
    p2 = (-gamma - sqrt(discriminant)) / 2;
    
    // Using a simpler, more stable basis spanning the same space
    B[,1] = exp(p1 * t_vector);
    B[,2] = exp(p2 * t_vector);

  } else if (discriminant < -1e-8) { // Case 3: Complex Roots
    alpha = -gamma / 2;
    omega = sqrt(-discriminant) / 2;
    
    B[,1] = exp(alpha * t_vector) * cos(omega * t_vector);
    B[,2] = exp(alpha * t_vector) * sin(omega * t_vector);
  
  } else { // Case 2: Repeated Real Roots
    p = -gamma / 2;
    
    B[,1] = exp(p * t_vector);
    B[,2] = t_vector * exp(p * t_vector);
  }
  
  return B;
}
``` -->

## Test functional form of FP

```{r, eval=FALSE}
library(mfp)
library(dplyr)
startx <- 1:80
d <- vroom::vroom('./data/pd.csv.bz2')
Mjs <- vroom::vroom(here('data', 'mjs.csv'))

dx = d %>% filter(sex == 'male', A == 'V', Z == 'M')

fit_mfp <- mfp(log(p_empirical) ~ fp(I(log(start)), scale = FALSE), data = dx)
y_pred_mfp <- predict(fit_mfp, newdata = tibble(start = startx))

# fit_mfp <- mfp(log(p_empirical+1e-12) ~ fp(I(log(start)), scale = FALSE), data = dD)
fit_mfp <- mfp(log(p_empirical+1e-12) ~ fp(((start)), scale = FALSE), data = dD)
fit_mfp <- mfp((p_empirical) ~ fp(((start)), scale = FALSE), data = dD)
y_pred_mfp <- predict(fit_mfp, newdata = tibble(start = startx))

library(TMB, help, pos = 2, lib.loc = NULL)
compile("notes/dfp2.cpp")
dyn.load(dynlib("code/dfp2"))

# Set up data and initial parameter list for TMB
data_list <- list(y = log(dx$p_empirical), x = dx$start)
param_list <- list(
  beta0 = 0, beta1 = 0, beta2 = 0,
  p1 = 0, p2 = 0,
  log_sigma = 0
)

# Create TMB objective function
obj <- MakeADFun(data = data_list, parameters = param_list, DLL = "dfp2", silent = FALSE)

# Optimize the model
opt <- nlminb(obj$env$last.par.best, obj$fn, obj$gr, control=list(maxit=1000))
opt
(rp <- obj$report())
# Get final parameter estimates and standard errors
(sd_report <- sdreport(obj))

x_pred_grid <- 1:80

# 1. Predict from the fitted DFP-2 model

A_fit <- matrix(c(0, -rp$delta, 1, -rp$gamma), 2, 2)
t_pred <- log(x_pred_grid)
basis_pred <- t(sapply(t_pred, function(t) {
  S <- expm::expm(A_fit * t)
  c(B1 = S[1, 1], B2 = S[1, 2])
}))
y_pred_dfp2 <- rp$beta0 + 
               rp$beta1 * basis_pred[, "B1"] + 
               rp$beta2 * basis_pred[, "B2"]
lines(x_pred_grid, (y_pred_dfp2), col = 4, lwd= 2)
plot(dx$start, dx$p_empirical, xlim = c(0, 80))
plot(log(dx$start), log(dx$p_empirical))
lines(log(x_pred_grid), (y_pred_dfp2), col = 2)
```